---
title: "Pokemon Winner Prediction"
author: "Ana Maria, Arturo, Shuaib, Viktor"
date: "1/9/2019"
output:
  html_document:
    toc: true
    theme: lumen
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

## 1 Introduction

### 1.1 The Data

```{r, include=FALSE}
setwd("~/Documents/Pokemon-winner-prediction")
pokemon <- read.csv('./pokemon.csv')
fights <- read.csv('./combats.csv')
```


X. ID for each pokemon
Name: Name of each pokemon
Type 1: Each pokemon has a type, this determines weakness/resistance to attacks
Type 2: Some pokemon are dual type and have 2
Total: sum of all stats that come after this, a general guide to how strong a pokemon is
HP: hit points, or health, defines how much damage a pokemon can withstand before fainting
Attack: the base modifier for normal attacks (eg. Scratch, Punch)
Defense: the base damage resistance against normal attacks
SP Atk: special attack, the base modifier for special attacks (e.g. fire blast, bubble beam)
SP Def: the base damage resistance against special attacks
Speed: determines which pokemon attacks first each round

```{r cars}
summary(cars)
```

## Data Exploration

```{r load, include=FALSE}
#Libraries
library(ggplot2)
library(dplyr)
library(gridExtra)
#install.packages('corrplot')
library(corrplot)
#install.packages('caret')
library(caret)
#install.packages('ggthemes')
library(ggthemes)
library(RColorBrewer)
#install.packages('fmsb')
library(fmsb)
#install.packages('rpart.plot')
library(rpart.plot)
#install.packages('ROCR')
library(ROCR)
library(ggplot2)
```
You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
